Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-15 probabilisticProgrammingMartingales.tex
Fix examples of partial antitone ranking functions.

\begin{example}
\label{ex:ac-ranking}
\begin{enumerate}
\item 1D unbiased random walk
\[
M_1 = 
\big(\tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
\]

\item 1D biased (away from 0) random walk
\[
M_2 = 
\big(\tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 1)}\big) \, 10
\]

\item 1D biased (towards 0) random walk
\[
M_3 = 
\big(\tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{2/3} f \, (n + 2)}\big) \, 10
\]

@@

\paragraph{Problem.} Find a generalised notion of ranking function 
%--- call it \emph{antitone-convex rankable} --- 
so that $M_1$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.

@@

The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) and convex function $\epsilon : \pReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
\[
f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
\]
where $R \to R’$ is the $\tY$-redex rule.

@@

\begin{definition}
We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ \akr{Separating the stopping time from $Y_n = 0$ allows for a neater definition of the ranking function of a term} \lo{Nice!} adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone measurable function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
\]
\end{definition}

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
Then $T < \infty$ almost surely.
\akr{It's not actually true that $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon_0}$ where $\epsilon_0 := \epsilon(\expect{Y_0})$. This is fortunate because it means that a rankability theorem based on antitone strict supermartingales actually has a chance of being stronger than the existing one (which was pretty much complete for $\tY$-PAST terms already). In fact, for any almost surely finite stopping time, there is a corresponding antitone strict supermartingale. This seems promising for the method being conditionally complete (for terms where all reachable terms are AST).}
\lo{See \Cref{sec:Relativised completeness}.}
\end{theorem}

@@

Programs $M_1$ and $M_3$ are AST.
To construct antitone ranking functions, we appeal to a partial ranking function theorem for antitone ranking functions.

For $M_1$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
Using shorthand 
$\Theta_1 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone partial ranking function $f_1 : \mathit{Rch}(M_1) \rightharpoonup \nnReal$ with antitone function $\epsilon_1$ defined as follows (for $n \in \mathbb N$):
\begin{align*}
{\Theta_1} \, n 
&\mapsto 
g_1(n)
\\
0 &\mapsto 0.
\end{align*}
For $n \geq 1$, $\Theta_1 \, n$ reduces in several steps to either $\Theta_1 \, (n-1)$ or $\Theta_1 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
\begin{align*}
  & g_1(n) - \frac{g_1(n-1) + g_1(n+1)} 2 \\
  = & \log(\frac{n+1}{\sqrt{n(n+2)}}) \\
  = & \frac 1 2 \log(\frac{(n+1)^2}{n(n+2)}) \\
  = & \frac 1 2 \log(1 + \frac 1 {n(n+2)}) \\
  > & \frac 1 {4n(n+2)} \\
  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
\end{align*}
and $\Theta_1 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_1$ of $1$, therefore by setting $\epsilon_1(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_1$ must decrease is met. It is also defined at $M_1 = \Theta_1 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_1$ is AST.

\begin{example}[Continuous random walk]\label{ex:raven complex}
In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
\[
\underbrace{\big
(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample))} \big)}_{\Theta} \, 10
\]
We can construct a partial antitone ranking function $f$ as follows:
\begin{align*}
\Theta \, l 
&\mapsto 
l
\\
\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
&\mapsto
l - 1/2
\\
0 &\mapsto 0
\end{align*}
with the constant antitone function $\epsilon(y) = 1/2$.

For a more complex example, consider the following ``continuous random walk''
\[
\underbrace{\big
(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x + \tsample)} \big)}_{\Xi} \, 10
\]
Let $g(x) := \sqrt{2} - \frac{1}{2^{x-\frac{1}{2}}}$ and let $\epsilon$ be the function specified by
\[
\epsilon(g(x)) = g(x) - \dfrac{1}{2}\Big(g(x - \dfrac{1}{2}) + \cdot g(x + \dfrac{1}{2}) \Big) = 
\dfrac{3 - 2\sqrt{2}}{2^{x+1}}.
\]
We define a partial antitone ranking function as follows:
\begin{align*}
\Xi \, l 
&\mapsto 
g(l)
\\
\tif{l \leq 0}{0}{\Xi \, (l - \tsample) \oplus_{1/2} \Xi \, (l + \tsample)}
&\mapsto
\frac{1}{2}\big(g(l - \frac{1}{2}) + \cdot g(l + \frac{1}{2}) \big)
\\
r \leq 0, \quad r &\mapsto 0
\end{align*}
\end{example}

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, 10
\]
To construct an antitone ranking function, we solve the recurrence relation:
\[
\left\{
\begin{array}{rll}
z_0 &=& 0\\
n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
\end{array}
\right.
\]
It turns out the solution to Example $M_1$, $z_n = \frac{2^n - 1}{2^{n-1}}$, is nearly a solution (for $n = 1$, we have equality instead).

Let $g(x) = \frac{2^n - 1}{2^{n-1}}$.
Then 
\begin{equation}
\epsilon(g(x)) = 
g(x) - \dfrac{x}{2x + 1} \, g(x-1) - \dfrac{x+1}{2x + 1} \, g(x+1) 
= \dfrac{x-1}{2^{x} \, (2x + 1)}
\label{eqn:in the limit}
\end{equation}

\lo{ongoing}

>>>>

\begin{example}
\label{ex:ac-ranking}
\begin{enumerate}
\item 1D biased (towards 0) random walk
\[
M_1 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
\]

\item 1D unbiased random walk
\[
M_2 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
\]

\item 1D biased (away from 0) random walk
\[
M_3 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
\]

@@

The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and Theorem \ref{thm:RankingSoundness} is insufficient to prove its termination. We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.

@@

The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
\[
f(E[R’]) \leq f(E[R]) - \epsilon(f(E[R])) 
\]
where $R \to R’$ is the $\tY$-redex rule.

@@

\begin{definition}
We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
\]
\end{definition}

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
Then $T < \infty$ almost surely.
\end{theorem}

@@

Programs $M_1$ and $M_2$ are AST. For $M_1$, the partial ranking function $\big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1$ suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).

For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
\begin{align*}
{\Theta_2} \, n 
&\mapsto 
g_2(n)
\\
0 &\mapsto 0.
\end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
\begin{align*}
  & g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  = & \log(\frac{n+1}{\sqrt{n(n+2)}}) \\
  = & \frac 1 2 \log(\frac{(n+1)^2}{n(n+2)}) \\
  = & \frac 1 2 \log(1 + \frac 1 {n(n+2)}) \\
  > & \frac 1 {4n(n+2)} \\
  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
\end{align*}
and $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$, therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.

\begin{example}[Continuous random walk]\label{ex:raven complex}
In SPCF we can construct a function whose argument changes by a random amount at each recursive call:
\[
\underbrace{\big
(\tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample))} \big)}_{\Theta} \, 10
\]
We can construct a partial ranking function $f$ as follows:
\begin{align*}
\Theta \, l 
&\mapsto 
l + 2
\\
\tif{l \leq 0}{0}{\Theta \, (l - \tsample)}
&\mapsto
l + 1
\\
0 &\mapsto 0.
\end{align*}

For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk''
\[
\underbrace{\big
(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} \big)}_{\Xi} \, 10
\]
Let $g(x) := 2 + log(x + 1)$ again and let $\epsilon$ be the function specified by
\[
\epsilon(g(x-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y.
\]
The limit of this as $x \to \infty$ and $g(x+1/2) \to \infty$ is 0, and $\frac {\mathrm d}{\mathrm dx} \epsilon(g(x+1/2)) = \frac 1 {x+1} + \log(1 - \frac 1 {x + 1/2}) < 0$, and $g$ is monotonic increasing therefore $\epsilon$ is antitone and bounded below by 0.
We define a partial antitone ranking function as follows:
\begin{align*}
\Xi \, l 
&\mapsto 
g(l)
\\
0 &\mapsto 0
\end{align*}
\end{example}
The value of $g$ after one $\tY$-reduction step is at least $g(l-1/2)$ therefore the expectation of $\epsilon$ after one $tY$-reduction step is at most $\epsilon(g(l-1/2)) = g(x) - \int_{x-1/2}^{x+1/2}g(y) \, \mathrm d y$, therefore in the case where $l > 0$, $g$ decreases by the required amount, and in the case where $l \leq 0$, $g(\Xi \, l) \geq 2 + log(1/2) > \log(\frac{3 \sqrt 3} 2 - 1 = \epsilon(g(0-1/2))$ as well, therefore this is a valid partial antitone ranking function and the term is AST.

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, 10
\]
To construct an antitone ranking function, we solve the recurrence relation:
\[
\left\{
\begin{array}{rll}
z_0 &=& 0\\
n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
\end{array}
\right.
\]
For $n > 2$, $z_n = \log(n-1)$ works. The expected decrease is $\frac 1 2(\log(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \log(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \log (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \log 2 - \log 3 - 1, z_1 = 3 \log 2 - 2 \log 3 - 3, z_0 = 4 \log 2 - 3 \log 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.
